A re-examination of the role of friction in the original Social Force Model

### 2019-06-21 paper.tex
correcciones de estilo

\textcolor{red}{In the last decades, many microscopic models 
for crowd dynamics were developed. The class of force-based model was introduced by Hirai 
and Tarui (1975) introducing the class of \textit{force based} models 
\cite{Hirai}. These models represent the interaction between 
pedestrians and their environment using different kinds of forces inspired in physical systems. For 
example, Okazaki (1979) modeled the movement of pedestrians as 
a magnetized object immersed a magnetic in field \cite{Okazaki}.}\\

By the late '90s and the beginning of the century, Helbing 
\textit{et al.} proposed a model that included socio-psychological and physical 
forces to simulate crowd dynamics. According to Helbing \textit{et al.}, both 
the environment and the individuals' own desire affect the pedestrians motion in 
a similar way as forces do with respect to the momentum of particles 
\cite{Helbing1,Helbing4}. This ``Social Force Model'' (SFM) related the 
socio-psychological phenomenon of crowds behavior to the ``microscopic'' 
formalism of moving particles. The model succeeded at this instance to 
reproduce the slow down in the efficiency in the evacuation process as 
pedestrians try harder to escape from a dangerous situation (\textit{i.e.} 
``faster is slower'' effect) \cite{Helbing1,Dorso1,Dorso2}. \\

@@

In the last years several extensions of the Social Force Model have been 
proposed. These extensions solve numerical pitfalls \cite{koster1} 
and other issues such as oscillations, overlapping and non-realistic trajectories
\cite{chraibi1,dietrich1}. However, many of these extensions surpass the 
crowded context of the original SFM \cite{Helbing1,Helbing4},focusing on the 
individual behavior of each agent. The basic SFM assumes strong interactions
in a dense environment.\\

Some questioning arose on the true psychological tendency of the pedestrians to 
stay away from each other. While the social forces accomplish this tendency, it 
attains a somewhat unrealistic ``colliding behavior'' for slowly moving 
pedestrians \cite{Lakoba}. His (her) repulsive tendency is expected to decrease 
as approaching a more crowded environment. The small fall-off length $B=0.08\,$m 
suggested by Helbing in Ref.~\cite{Helbing1} does not completely solve this 
issue. It neither agrees with the fact that pedestrians prefer to keep a 
comfortable $0.5\,$m distance between each other in a moderately crowded 
environment, nor it fits accurately the empirical velocities reported for 
crowds under normal conditions. \cite{Lakoba}.  \\

@@

\textcolor{red}{More recent findings on the fundamental diagram for an extremely dense
 situation show that, sometimes, the flux may increase with the density (see Ref.\cite{lohner1}). 
This seems to contradict the results of Ref.~\cite{helbing3}. But both results appear 
not to be completely  comparable since data was acquired at different points, and thus,
at different stages of the ritual. Ref.~\cite{lohner1}, indeed, states that the fundamental 
diagram may strongly depend on the type of motion and the nature of the event.}\\

\textcolor{red}{ Our work focuses on data from Ref.~\cite{helbing3}. 
Further measurements can be found in Refs.~\cite{Boltes,Dridi1,Dridi2,Baqui1,lohner1}.
The PedFlow model developed by L\"ohner et al. Ref.~\cite{lohner2}  
is an alternative to the Helbing model. Togashi et al.  Ref.~\cite{Togashi1} 
improved the calibration of this model using a Kalman filter ensemble. }\\

@@

The investigation is organized as follows. We first recall the SFM in 
Section~\ref{sfm}, while including the precise definitions for flux, density 
and clustered structures. Section~\ref{simulations} presents our numerical 
simulations for pedestrians moving through corridors. The corresponding results 
are shown in Section~\ref{results}. Our main conclusions are detailed in the 
closing Section~\ref{conclusions}.\\

@@

where the subscript $j$ represents all the other pedestrians 
(excluding $i$).\\

@@

The simulations process were performed on a straight corridor of length $L=$ 
28~m (with periodic boundary conditions) and variable width $w$. We explored 
widths ranging from $w=2$~m to $w=40$~m. The corridor had two side walls, placed 
at $y=0$ and $y=w$, respectively. The length of each wall was $L$. The 
pedestrians were modeled as soft spheres of radius $R_i=0.23$~m. This size was 
fixed according to Ref.~\cite{metric_handbook}. Initially, the individuals were 
randomly distributed along the corridor with a fixed global density 
(\textcolor{red}{p~m$^{-2}$}) and with random initial velocities, resembling a 
Gaussian distribution with null mean value. We explored global density values in 
the range 1\textcolor{red}{p~m$^{-2}$}~$<\rho<$~9~\textcolor{red}{p~m$^{-2}$}. 
We did not explore extreme densities (say $\rho>$9~\textcolor{red}{p~m$^{-2}$}) 
because we exclude injuring situations. The number of pedestrians in the 
simulation was given by the global density and the corridor dimensions chosen in 
each case. \\

@@

In the case of narrow corridors, both the simulated case and the empirical results converge to a constant flow value. It is remarkable that the system does not reach a freezing state such as the one reported in Refs.\cite{kwak,lin}. Recall that our simulations do not include any respect factor (see Ref.\cite{parisi2}), or changes in the net-time headway (Ref.\cite{helbing3}), or the urge to see an attraction (Ref.\cite{kwak}). We assume a well defined target and the same $v_d$ for all the pedestrians.\\

The inset in Fig.~\ref{fundamental_diagram_flow} corresponds to the empirical data from Helbing (\cite{helbing3}). at the entrance of the Jamaraat bridge (the corridor width was $w=22$~m). Notice that our simulated results corresponding to a $w=22$~m corridor, exhibit a different behavior along the congested regime. In the simulated case, the flow increases even for the greatest explored density. On the contrary, the empirical data exhibit a flow reduction for $\rho > 5$ until reaching a plateau for the highest explored density values.  \\

@@

We want to mention that there is a clear relation between $v_{max}$ and the corridor width. That is, the wider the corridor, the higher the maximum reached velocity. (Fig.~\ref{speed-profile-width-normaliz} does not exhibit this behavior because the velocity is normalized by $v_{max}$).\\

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./fracc_clusteriz_vs_density.eps}
\caption{\label{fic} Fraction of clustered individuals as a function of the density. Squared symbols correspond to $\kappa_w=\kappa_i=2.4\times 10^6$, circles to $\kappa_i=2.4\times 10^6$ and $\kappa_w=2.4\times 10^5$, triangles to to $\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^6$ and crosses to the coefficients of the original Social Force Model ($\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^5$). The pedestrians walk across a corridor with $v_d=1$m~s$^{-1}$. The measurements were recorded every 0.5~s once the system reached the stationary state. The values corresponding to the fraction of clustered individuals were averaged over 170~s. The cluster cutoff distance was 0.46~m (equivalent to the shoulder width of the pedestrians).}
\end{figure}

@@

Our investigation focused on the fundamental diagram in the context of the 
social force model. We compared empirical data recorded at 
the entrance of the Jamaraat bridge (see Ref.~\cite{helbing3}) with our own SFM 
simulations. We observed  that the  social force model in its 
original version does not fit into the empirical fundamental 
diagram since the pedestrian flow increases even for high dense 
crowds. The reasons for this mismatching were studied through 
numerical computations and by a simple theoretical example. We arrived to the 
conclusion that either increasing the friction coefficient or increasing the 
relaxation time may be the key for achieving a non-increasing 
flow in the congested regime of the fundamental diagram. The 
latter has already been explored in Ref.~\cite{johansson} and a 
similar idea was introduced in Ref.\cite{parisi2}. We noticed, 
though, that both approaches are equivalent since both affect 
the reduced-in-units equation of motion in a similar fashion.\\

@@

In this Appendix we show that the friction modification does not alter already studied behaviors of pedestrian dynamics (see Refs.~\cite{Dorso1,Dorso2,Dorso3}). In order to check this out, we performed numerical simulations of a room of 225 pedestrian escaping through a narrow door (bottleneck enclosure). The evacuation time (clearance time) $t_e$ is a function of the desired velocity $v_d$. This function has a minimum such that below it, $t_e$ is a decreasing function of $v_d$ while above, the tendency is reversed. It means that the evacuation process is optimum at moderated $v_d$ . This is a well known phenomenon called the Faster-is-Slower effect. This effect was reported in numerical simulations \cite{Dorso1,Dorso5} and it has also experimental support \cite{parisi1}. \\

In Fig.~\ref{fis} we show the evacuation time as a function of the desired velocity corresponding to a friction coefficient ten times grater than the friction coefficient of the original Social Force Model (\textit{i.e.} now $\kappa=2.4\times 10^{6}$). We can see that this modification preserves the Faster-is-Slower effect (and the Faster-is-faster effect). The inset exhibits the evacuation time as a function of the desired velocity with the friction coefficient of the original model ($\kappa=2.4\times 10^{5}$).\\

@@

Fig.~\ref{rate_fis} exhibits the ratio between the evacuation time with for the friction modified model and the evacuation time for the original model. This ratio is shown as a function of the desired velocity. As the desired velocity increases, the rate of evacuation times approaches a constant value $\sim$ 9. This result is in agreement with the formula Eq.(9) in Ref.\cite{sticco}. The friction force dominates in the range of high desired velocities since it promotes the collisions between pedestrians. within this regime, the Eq. (9) of  Ref. \cite{sticco} becomes a good approximation of the evacuation time. Thus, it is reasonable that dividing $t_e$ (for the enhanced friction) by $t_e$ (for the original model) yields a constant value for high $v_d$.

@@

 \bibitem{Okazaki}
\textcolor{red}{Okazaki, S., 1979. A study of pedestrian movement in 
architectural space, Part 
1: pedestrian movement by the application of magnetic model. Trans. A.I.J. 283, 
111–119.}

>>>>

\textcolor{red}{In the last decades, many microscopic models 
for crowd dynamics were developed. The class of force-based model was proposed by Hirai 
and Tarui (1975) introducing the class of \textit{force based} models 
\cite{Hirai}. These models represent the interaction between 
pedestrians and their environment using different kinds of forces inspired in physical systems. For 
example, Okazaki (1979) modeled the movement of pedestrians as 
a magnetized object immersed a magnetic in field \cite{Okazaki}.}\\

By the late '90s and the beginning of the century, Helbing 
\textit{et al.} proposed a model that included socio-psychological and physical 
forces to simulate crowd dynamics. According to Helbing \textit{et al.}, both 
the environment and the individuals' own desire affect the pedestrians motion in 
a similar way as forces do with respect to the momentum of particles 
\cite{Helbing1,Helbing4}. This ``Social Force Model'' (SFM) related the 
socio-psychological phenomenon of crowds behavior to the ``microscopic'' 
formalism of moving particles. The model succeeded at this instance to 
reproduce the slow down in the efficiency in the evacuation process as 
pedestrians try harder to escape from a dangerous situation (\textit{i.e.} 
``faster is slower'' effect) \cite{Helbing1,Dorso2}.
\textcolor{red}{The faster-is-slower effect was well explained by the presence of human blocking clusters 
that obstruct the exit during an emergency evacuation \cite{Dorso1}.}  \\

@@

In the last years several extensions of the Social Force Model have been 
proposed. These extensions solve numerical pitfalls \cite{koster1} 
and other issues such as oscillations, overlapping and non-realistic trajectories
\cite{chraibi1,dietrich1}. However, many of these extensions surpass the 
crowded context of the original SFM \cite{Helbing1,Helbing4}, focusing on the 
individual behavior of each agent. The basic SFM assumes strong interactions
in a dense environment.\\

Some questioning arose on the true psychological tendency of the pedestrians to 
stay away from each other. While the social forces accomplish this tendency, it 
attains a somewhat unrealistic ``colliding behavior'' for slowly moving 
pedestrians \cite{Lakoba}. His (her) repulsive tendency is expected to decrease 
as approaching a more crowded environment. The small fall-off length $B=0.08\,$m 
suggested by Helbing in Ref.~\cite{Helbing1} does not completely solve this 
issue. It neither agrees with the fact that pedestrians prefer to keep a 
comfortable $0.5\,$m distance between each other in a moderately crowded 
environment, nor it fits accurately the empirical velocities reported for 
crowds under normal conditions \cite{Lakoba}.  \\

@@

\textcolor{red}{More recent findings on the fundamental diagram for an extremely dense
 situation show that, sometimes, the flux may increase with the density (see Ref.~\cite{lohner1}). 
This seems to contradict the results of Ref.~\cite{helbing3}. But both results appear 
not to be completely  comparable since data was acquired at different points, and thus,
at different stages of the ritual. Ref.~\cite{lohner1}, indeed, states that the fundamental 
diagram may strongly depend on the type of motion and the nature of the event.}\\

\textcolor{red}{ Our work focuses on data from Ref.~\cite{helbing3}. 
Further measurements can be found in Refs.~\cite{Boltes,Dridi1,Dridi2,Baqui1,lohner1}.
The PedFlow model developed by L\"ohner \textit{et al}. Ref.~\cite{lohner2}  
is an alternative to the Helbing model. Togashi \textit{et al}.  Ref.~\cite{Togashi1} 
improved the calibration of this model using a Kalman filter ensemble. }\\

@@

The investigation is organized as follows. We first recall the SFM in 
Section~\ref{sfm}, while including the precise definitions for flux, density 
and clustered structures. Section~\ref{simulations} presents our numerical 
simulations for pedestrians moving through corridors. The hypotheses of the 
work are mentioned in Section~\ref{Hypotheses}. The corresponding results 
are shown in Section~\ref{results}. Our main conclusions are detailed in the 
closing Section~\ref{conclusions}.\\

@@

where the subscript $j$ represents all the other pedestrians or walls
(excluding pedestrian $i$).\\

@@

The simulations process were performed on a straight corridor of length $L=$ 
28~m (with periodic boundary conditions) and variable width $w$. We explored 
widths ranging from $w=2$~m to $w=40$~m. The corridor had two side walls, placed 
at $y=0$ and $y=w$, respectively. The length of each wall was $L$. The 
pedestrians were modeled as soft spheres of radius $R_i=0.23$~m. This size was 
fixed according to Ref.~\cite{metric_handbook}. Initially, the individuals were 
randomly distributed along the corridor with a fixed global density 
(\textcolor{red}{p~m$^{-2}$}) and with random initial velocities, resembling a 
Gaussian distribution with null mean value. We explored global density values in 
the range 1~\textcolor{red}{p~m$^{-2}$}~$<\rho<$~9~\textcolor{red}{p~m$^{-2}$}. 
We did not explore extreme densities (say $\rho>$9~\textcolor{red}{p~m$^{-2}$}) 
because we excluded injuring situations. The number of pedestrians in the 
simulation was given by the global density and the corridor dimensions chosen in 
each case. \\

@@

In the case of narrow corridors, both the simulated case and the empirical results converge to a constant flow value. It is remarkable that the system does not reach a freezing state such as the one reported in Refs.~\cite{kwak,lin}. Recall that our simulations do not include any respect factor (see Ref.~\cite{parisi2}), or changes in the net-time headway (Ref.~\cite{helbing3}), or the urge to see an attraction (Ref.~\cite{kwak}). We assume a well defined target and the same $v_d$ for all the pedestrians.\\

The inset in Fig.~\ref{fundamental_diagram_flow} corresponds to the empirical data from Helbing (see Ref.~\cite{helbing3}). at the entrance of the Jamaraat bridge (the corridor width was $w=22$~m). Notice that our simulated results corresponding to a $w=22$~m corridor, exhibit a different behavior along the congested regime. In the simulated case, the flow increases even for the greatest explored density. On the contrary, the empirical data exhibit a flow reduction for $\rho > 5$ until reaching a plateau for the highest explored density values.  \\

@@

We want to mention that there is a clear relation between $v_{max}$ and the corridor width. That is, the wider the corridor, the higher the maximum reached velocity (Fig.~\ref{speed-profile-width-normaliz} does not exhibit this behavior because the velocity is normalized by $v_{max}$).\\

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./fracc_clusteriz_vs_density.eps}
\caption{\label{fic} Fraction of clustered individuals as a function of the density. Squared symbols correspond to $\kappa_w=\kappa_i=2.4\times 10^6$, circles to $\kappa_i=2.4\times 10^6$ and $\kappa_w=2.4\times 10^5$, triangles to $\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^6$ and crosses to the coefficients of the original Social Force Model ($\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^5$). The pedestrians walk across a corridor with $v_d=1$m~s$^{-1}$. The measurements were recorded every 0.5~s once the system reached the stationary state. The values corresponding to the fraction of clustered individuals were averaged over 170~s. The cluster cutoff distance was 0.46~m (equivalent to the shoulder width of the pedestrians).}
\end{figure}

@@

Our investigation focused on the fundamental diagram in the context of the 
Social Force Model. We compared empirical data recorded at 
the entrance of the Jamaraat bridge (see Ref.~\cite{helbing3}) with our own SFM 
simulations. We observed  that the  social force model in its 
original version does not fit into the empirical fundamental 
diagram since the pedestrian flow increases even for high dense 
crowds. The reasons for this mismatching were studied through 
numerical computations and by a simple theoretical example. We arrived to the 
conclusion that either increasing the friction coefficient or increasing the 
relaxation time may be the key for achieving a non-increasing 
flow in the congested regime of the fundamental diagram. The 
latter has already been explored in Ref.~\cite{johansson} and a 
similar idea was introduced in Ref.~\cite{parisi2}. We noticed, 
though, that both approaches are equivalent since both affect 
the reduced-in-units equation of motion in a similar fashion.\\

@@

In this Appendix we show that the friction modification does not alter already studied behaviors of pedestrian dynamics (see Refs.~\cite{Dorso1,Dorso2,Dorso3}). In order to check this out, we performed numerical simulations of a room of 225 pedestrian escaping through a narrow door (bottleneck enclosure). The evacuation time (clearance time) $t_e$ is a function of the desired velocity $v_d$. This function has a minimum such that below it, $t_e$ is a decreasing function of $v_d$ while above, the tendency is reversed. It means that the evacuation process is optimum at moderated $v_d$. This is a well known phenomenon called the Faster-is-Slower effect. This effect was reported in numerical simulations \cite{Dorso1,Dorso5} and it has also experimental support \cite{parisi1}. \\

In Fig.~\ref{fis} we show the evacuation time as a function of the desired velocity corresponding to a friction coefficient ten times grater than the friction coefficient of the original Social Force Model (\textit{i.e.} now $\kappa=2.4\times 10^{6}$). We can see that this modification preserves the Faster-is-Slower effect (and the Faster-is-Faster effect). The inset exhibits the evacuation time as a function of the desired velocity with the friction coefficient of the original model ($\kappa=2.4\times 10^{5}$).\\

@@

Fig.~\ref{rate_fis} exhibits the ratio between the evacuation time for the friction modified model and the evacuation time for the original model. This ratio is shown as a function of the desired velocity. As the desired velocity increases, the rate of evacuation times approaches a constant value $\sim$ 9. This result is in agreement with the formula Eq.~(9) in Ref.~\cite{sticco}. The friction force dominates in the range of high desired velocities since it promotes the collisions between pedestrians. within this regime, the Eq. (9) of  Ref.~\cite{sticco} becomes a good approximation of the evacuation time. Thus, it is reasonable that dividing $t_e$ (for the enhanced friction) by $t_e$ (for the original model) yields a constant value for high $v_d$.

@@

 \bibitem[Okazaki (1979)]{Okazaki}
\textcolor{red}{Okazaki, S., 1979. A study of pedestrian movement in 
architectural space, Part 
1: pedestrian movement by the application of magnetic model. Trans. A.I.J. 283, 
111–119.}
